---
title: "A5: Incarceration"
output: html_document
---

### Introduction

The US prison system has many issues, with one of the most important ones being the over-incarceration of POC. The Vera Institute of technology, from where this data is derived from, aims to fight this over-incarceration and bring justice to the people from the unfair prison system altogether.

The variables this report will analyze will be:

(1) the change in races and sexes of prisoners in all counties over time in order to see how the race and sex compositions of prisoners over WA state have changed over time

(2) the relationship between rurality of a county and the race composition of prisoners in that county to determine if there's a correlation between rural counties and prisoner compositions

(3) the distribution of race and sexes of prisoners in a county on a map for the most recent date to see a clearer concentration of what compositions of prisoners are in each county today and whether there is still inequality

### Summary Information

Some information found on the dataset can be seen by the summary statistics as shown below. The first summary stat finds the county with the highest jail population rate. It is Yakima County at a jail pop rate of 992.82. The county with the lowest jail population rate is Whitman County at 81.31 prisoners. The highest number of recorded POC prisoners was 8215.3 and the county with the highest number of POC prisoners is Adams County with 4255 prisoners. The year with the highest prisoners was 2009 where Yakima county had a total jail pop rate of 922.82.

``````{r, echo = FALSE, code = readLines("summary.R")}
```

### The Dataset

The data was collected by the Vera Institute of Justice.

The data was collected in various ways by the Vera Institute. Some ways are from the government and government agencies, criminal settings such as jails and courtrooms, city/town halls. This Washington dataset may have been retrieved using data from the WA state government and similar areas.

The data was collected in order to bring about ending mass incarceration and searching for solutions to do that. The Vera Institute wants to transform the justice system and they can do so by studying prison populations and demographics of who is in prison to determine if there's bias towards particular groups of people.

```{r}
print(nrow(wa_data))
```

There are 1131 rows in this dataset.

How many features (columns) are in the data? 

```{r}
print(ncol(wa_data))
```

There are 23 columns in this dataset.

The ethical concerns with this data are that since this data was gathered by the Vera Institute from various sources, it is unknown if there is under-reporting or over-reporting of poeple who are actually in prisons during specific times. This can cause issues as it may not accurately represent number of POC in prisons which is what our analysis is trying to cover.

The limitations are that this data only has the races of people who are in prison in WA state, and it's a small amount of data compared to the entire set. The data also does not have the reasonings behind why prisoners are in prison. We won't be able to know the severity of crimes and peoples' sentences in prison, which may or may not play a role in determining why the demographics of prisoners is the way it is. So essentially, the dataset is generalized and we do not know the specifics. It also has a lot of N/A columns for population of specific races in prisons. This further adds to the issues and questions of how accurate the demographics are. It also only sorts the races into male and female, aapi, black, latinx, native, and white, so we have to consider that while the races and sexes are generally accounted for, it is unsure how people who are a mix of these may be represented in this data. Some counties will smaller populations may also not have enough representation compared to counties with larger populations. Also, if we are looking at this for only Washington state, it raises a lot of questions for how the data will be represented in other states' datasets too.

### Trends Over Time Chart

This chart shows the jail population rates of various racial groups in King County jails from the year 1990 to 2018. This chart was included in order to research any over-incarceration of races in King County and whether or not that has changed over time. The results are really interesting, as it shows black people were over-incarcerated in the 1990s, but that has steadily decrease over time to 2018, showing that while they are still over-incarcerated, the amount of black people in prison has decreased overall. Latinx people had a spike in incarcerations in 1997 but that has also decreased. Overall though, POC have higher rates of incarceration than white people.

``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The second chart that you will create and include will show how two different (continuous) variables are related to one another. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- You must have clear x and y axis labels
- The chart needs a clear title 
- If you choose to add a color encoding (not required), you need a legend for your different color and a clear legend title

``````{r, echo = FALSE, code = readLines("chart2_example.R")}
```

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

``````{r, echo = FALSE, code = readLines("map_example.R")}
```

